RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2014-11-21 manuscripts/manuscript.Rmd
Resolved a number of comments during conference call.

<!-- Map paragraph -->

@@

A NeXML document can be thought of as having the following components: (1) phylogeny
data, (2) character-trait data, (3) operational taxonomic units, and (4) metadata. 
`RNeXML` seeks to provide a seamless way to convert these to and from their 
corresponding R object representations, in particular `ape::phylo`,
`data.frame`, and lists for phylogenies, character
matrices, and metadata, respectively.
<!-- I feel the following is a little early - it will likely be over -->
<!-- most most readers' heads. HL
(or for advanced users,
`XMLInternalDocument` and `rrdf` triplestores).
-->
Metadata can provide bibliographic and provenance
information about the study as part of which the phylogeny was generated or
applied (such as authors, citations to publications, timestamps, and even
descriptions of methods). They can also be used to attach annotations to individual
data elements.  We illustrate how metadata annotations can be added to
the operational taxonomic units of the phylogeny and can be used to extend
the data representation to new uses such as stochastic character maps.
<!-- sounds awkward - need to smoothen the language. HL -->

@@

```{r}
library("RNeXML")
nex <- nexml_read("components/trees.xml")
```
The resulting `nex` object is of type `nexml` and is a native R object representation of the NeXML
file. <!-- This seems redundant to me to what has already been 
said. HL -->
<!-- CB: Tried to revise what follows to emphasize that we have a new data structure / object class here,
that has unique merits -->
Objects of type `nexml` represent phylogenetic trees as well as 
character traits, and semantic metadata. As such, this new R object format differs
from existing formats such as `ape::phylo` format [@Paradis2004],
which describes trees alone, or any of the `phylobase` classes of object that contain 
trees and traits but more limited metadata [@phylobase].

@@

Reading phylogenies from TreeBASE into R was already possible using
the R package `treebase` [@Boettiger_2012]. However, because it relied
on the NEXUS parser provided in the `ape` package, and because of
incompatible flavors of the NEXUS standard [@Vos_2012], 
a significant number <!-- State explicitly --> of TreeBASE NEXUS
files could not be read in this way.  Owing to the syntactically strict
definition of the
NeXML standard, the RNeXML package can successfully read every 
NeXML document from TreeBASE. <!-- are there invalid ones as well that -->
<!-- it can't read? the sentence as is suggests so, and so is a little -->
<!-- awkward (it leaves open the possibility that the NEXUS files ape -->
<!-- can't read are now invalid NeXML files. Which is probably not -->
<!-- what you mean to suggest. HL -->
<!-- Correct, there are indeed a few invalid ones (e.g. some are purely
erroneous HTML, see issues on supertreebase repo) which it cannot
parse.  But like you say, no need to confuse the reader.  I've dropped the 
word 'valid'.  -->

@@

```{r }
nex <- nexml_read("components/comp_analysis.xml")
get_characters(nex)
```
<!-- same as above - why can this not be one call? HL -->
<!-- CB: I've dropped the system.file, even though that is sadly the 
standard way for R to load data that comes with a package but is not in CSV, R, or .rda format.

Indeed `get_characters` could call the file path directly, but there is key
conceptual reason for not doing this.  Note that all the examples call nexml_read to get a nex object
before using get_characters, get_trees, etc.  That is because we really want the user to get used
to manipulating the nexml R data structure, rather than using the external XML file directly.

The functions embody this design principle deep down.  We could have written get_characters etc
in a way that it just yanked the characters out of the XML and left everything else, rather than 
converting the whole NeXML file into a serialized R object stored in memory.  But the latter strategy
gives us a native R object that people can extend, write functions for, and subset and query with 
the standard R syntax.  Without this, the package isn't really providing a native format, and users 
and developers are left on their own to either also work directly with the XML or just convert immediately
into a more limited R format like ape::phylo.  
-->

Continuous and discrete character data are represented as separate
blocks in NeXML files, but will be combined as separate columns of a
single data frame by `get_characters()` if they correspond to the same
taxa. If this behavior is undesired, the function
`get_characters_list()` can be used to obtain
a list of data frames, where list elements correspond to the character matrix blocks of
the NeXML file.

@@

Here we create a `nexml` object for the `bird.orders` data
provided in the `ape` package, and then add basic metadata:
<!-- I notice that other add_* methods take the nexml object as the -->
<!-- last argument, not the first as seems to be the case here. This -->
<!-- is inconsistent and should be fixed. HL -->

<!-- Not so, the nexml object is the second argument here too.  If no existing 
nexml object is given, the add function starts a new nexml object.  If one is
given, it appends it's first argument onto the existing nexml, as described above.

Notice that bird.orders is an ape::phylo object, which we are adding to a new 
nexml object.  (R allows for default arguments, the default second argument is 
new("nexml") -->

@@

  creator = "Sibley, C. G. and Ahlquist, J. E.")
```

>>>>

@@

A NeXML document can be thought of as having the following components: (1) phylogeny
data, (2) character-trait data, (3) operational taxonomic units, and (4) metadata. 
`RNeXML` seeks to provide a seamless way to convert these to and from their 
corresponding R object representations, in particular `ape::phylo`,
`data.frame`, and lists for phylogenies, character
matrices, and metadata, respectively.
Metadata can provide bibliographic and provenance
information about the study as part of which the phylogeny was generated or
applied (such as authors, citations to publications, timestamps, and even
descriptions of methods). They can also be used to attach annotations to individual
data elements.  We illustrate how metadata annotations can be added to
the operational taxonomic units of the phylogeny and can be used to extend
the data representation to new uses such as stochastic character maps.
<!-- sounds awkward - need to smoothen the language. HL -->

@@

```{r}
library("RNeXML")
nex <- nexml_read("components/trees.xml")
```
The resulting `nex` object is of type `nexml` and is a native R object representation of the NeXML
file. 
Objects of type `nexml` represent phylogenetic trees as well as 
character traits, and semantic metadata. As such, this new R object format differs
from existing formats such as `ape::phylo` format [@Paradis2004],
which describes trees alone, or any of the `phylobase` classes of object that contain 
trees and traits but more limited metadata [@phylobase].

@@

Reading phylogenies from TreeBASE into R was already possible using
the R package `treebase` [@Boettiger_2012]. However, because it relied
on the NEXUS parser provided in the `ape` package, and because of
incompatible flavors of the NEXUS standard [@Vos_2012], 
a sizable number of TreeBASE NEXUS
files could not be read in this way <!-- CB to rephrase to make a -->
<!-- broader point and refer back to arguments made in the NeXML -->
<!-- paper, possibly moving this larger point to the introduction -->.  Owing to the syntactically strict
definition of the
NeXML standard, the RNeXML package can successfully read every 
NeXML document from TreeBASE.

@@

```{r }
nex <- nexml_read("components/comp_analysis.xml")
get_characters(nex)
```
Continuous and discrete character data are represented as separate
blocks in NeXML files, but will be combined as separate columns of a
single data frame by `get_characters()` if they correspond to the same
taxa. If this behavior is undesired, the function
`get_characters_list()` can be used to obtain
a list of data frames, where list elements correspond to the character matrix blocks of
the NeXML file.

@@

Here we create a `nexml` object for the `bird.orders` data
provided in the `ape` package, and then add basic metadata:

@@

  creator = "Sibley, C. G. and Ahlquist, J. E.")
```
<!-- the above will get fixed by CB, this is just a reminder and -->
<!-- placeholder to adjust when that's done. -->
